Stateless Model Checking with Data-Race Preemption Points

### 2015-10-10 pldi-16/di.tex
talk about dr pps

@@

% TODO: Talk about the multiple ways DR PPs are added (alone / in addition)

% TODO: make sure to talk about avoiding thrashing
\subsection{Avoiding thrashing in \quicksand}

(Here the reviewers will wonder if the whole point of the paper was just to build up to that one joke.)

>>>>

% TODO: Overview

\subsection{Managing multiple state spaces}

@@

{\bf Avoiding thrashing in \quicksand.} % TODO

% TODO: decide which subsection should come first for better flow

\subsection{Data-race preemption points}

Intuitively, % TODO describe what we're trying to achieve

When \landslide~detects a data race, it produces a report for each of the two memory accesses involved in the race.
Each report indicates the program counter value (PC) associated with the access, as well as some further conditions to help filter away unrelated executions of the same instruction on different data.
(For example, {\tt list\_insert()} might be called from many parts of a codebase, but produce data-race candidates from only one callsite.)
Ideally, the PC would be qualified by a full backtrace, but tracing the stack is too expensive to do as often as each shared memory access.
Instead, \landslide~qualifies the PC with
(a) the current thread ID and
(b) the most recent {\tt call} instruction.
% (a crude approximation of a stack trace)
% which are much cheaper, as we carry them around all the time already
Note that we do {\em not} qualify data races by {\em memory addresses},
which change across thread interleavings.
% especially when malloc is involved.
Figure~\ref{fig:dont-filter-dr-by-address} shows example code where qualifying by memory address will miss the bug.

% TODO: make dont-filter-dr-by-address example.

When \quicksand~receives a data race report, it adds two new jobs to its workqueue:
a ``small'' job to preempt on the racing instruction only,
and a ``big'' job to preempt on that instruction as well as each PP used by the reporting job.
%
Hence each {\em pair} of racing accesses will spawn four new jobs, as shown in Figure~\ref{fig:new-dr-jobs}.
Rather than adding a maximal job with both new PPs at once, we prefer to add smaller jobs which have a higher chance of completing in time.
If those state spaces are bug-free, they will in turn add the maximal job later.
%
The rationale of spawning multiple jobs is that which will be more fruitful cannot be known in advance: while the big job may not complete in time, the small job might miss the data race entirely if the original PPs were required to expose it.
In practice, we have observed some bugs found quickly by the small job, and other bugs missed by the small job found eventually by the big job.
% which justifies this design

% TODO: make new-dr-jobs diagram

%                                  (2)          landslide         landslide
%  landslide                   "Preempt when-->  o  PPs=[foo]      o  PPs=[mutex_lock,
%    o  PPs=[mutex_lock]        PC = FOO"       / \               / \      foo]
%   / \                              |         o   o             o   o
%  o  /\                             |                          /|\  |\
%    o  o                                                      o o o o o
%                             __QUICKSAND__
%    |       (1)             |             |           (3)
%    \  "Hey, an access      |  job1       | ----"Preempt when -->
%     '- at FOO raced w/ --> |        job2 |      PC = BAR"
%        one at BAR!         |    job4     |
%                            |        job70|      [ same pictures for s/foo/bar/]
%                            | job8        |
%                            '-------------'
%                maybe show some internal structore of QS as well

### 2015-10-10 pldi-16/evaluation.tex
talk about dr pps

Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone in other model checkers.
For example, a single-state-space tool could use data-race candidates during immediately subsequent interleavings, essentially changing the state space on the fly.
Likewise, a message-passing-only tool could employ Iterative Deepening despite data races being absent from its concurrency model.
Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
we also sought to evaluate each technique individually.
Our evaluation answers the following questions:
\begin{enumerate}
	\item Is combining data-race analysis with stateless model checking productive?
		\begin{enumerate}
			\item Do data-race PPs expose new bugs that couldn't be found with single-state-space model checking?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
			\item Do we avoid false positives compared to single-pass?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
			% TODO: This one's optional. You can give up on it.
			\item Do we avoid false negatives compared to single-pass?
				% Explain later as: How many data-race bugs required backtracking to be identified as candidates in the first place
				%(don't forget set EXPLORE_BACKWARDS=0 here)
		\end{enumerate}
	\item Is Iterative Deepening better than the single-state-space model?
		\begin{enumerate}
			\item Does Iterative Deepening find bugs
				% in subset state spaces
				faster than the single-state-space approach, even without data races?
			\item Does Iterative Deepening help test data-race PPs?
				% Are there some DRs where the PP alone finds the bug, while the superset SS cannot, and some vice versa?
				% What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
		\end{enumerate}
\end{enumerate}
\subsection{Test Suite}
% TODO: Maybe say how many lines of code total? How many lines on average per P2/pintos? (careful with pintos; lots of basecode)
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
%across the Spring and Fall 2014 and Spring 2015 semesters;
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condvars, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtal memory implementations
\cite{pintos}.

@@

>>>>

Although \quicksand~presents Iterative Deepening and data-race PPs as interconnected techniques, they each could theoretically be employed alone in other model checkers.
For example, a single-state-space tool could use data-race candidates during immediately subsequent interleavings, essentially changing the state space on the fly.
Likewise, a message-passing-only tool could employ Iterative Deepening despite data races being absent from its concurrency model.
Hence, though many of our experiments compare \quicksand~to the state-of-the-art as a whole,
we also sought to evaluate each technique individually.
Our evaluation answers the following questions:
\begin{enumerate}
	\item Is combining data-race analysis with stateless model checking productive?
		\begin{enumerate}
			\item Do data-race PPs expose new bugs that couldn't be found with single-state-space model checking?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
			\item Do we avoid false positives compared to single-pass?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
			% TODO: This one's optional. You can give up on it.
			\item Do we avoid false negatives compared to single-pass?
				% Explain later as: How many data-race bugs required backtracking to be identified as candidates in the first place
				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
				% TODO: And disable false-neg malloc-free technique
				% TODO: And also disable the confirmed/suspected thing
				% (where mem.c waits for reorder observed before
				% messaging the latter half of the DR to QS)
		\end{enumerate}
	\item Is Iterative Deepening better than the single-state-space model?
		\begin{enumerate}
			\item Does Iterative Deepening find bugs
				% in subset state spaces
				faster than the single-state-space approach, even without data races?
			\item Does Iterative Deepening help test data-race PPs?
				% Are there some DRs where the PP alone finds the bug, while the superset SS cannot, and some vice versa?
				% What we want is a *variety*, which proves that iterative deepening is better than either individual strategy.
		\end{enumerate}
\end{enumerate}
\subsection{Test Suite}
% TODO: Maybe say how many lines of code total? How many lines on average per P2/pintos? (careful with pintos; lots of basecode)
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
%across the Spring and Fall 2014 and Spring 2015 semesters;
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condvars, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtal memory implementations
\cite{pintos}.

@@

In this section we show that using data-race PPs with \landslide~is more effective than either SSS-MC or single-pass data-race detection alone.

{\bf Finding bugs missed by SSS-MC.}
% TODO describe

{\bf Avoiding false positive data-race candidates.}
% TODO describe

{\bf Avoiding false negative data-race bugs.}
Some pairs of unprotected shared-memory accesses may be hidden in some control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass data-race detector
%could not achieve the coverage necessary
would fail
to identify the access pair as a data-race candidate to begin with.
%
We counted how many such data-races, used as PPs, would lead to \landslide~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
% TODO: Put a figure here giving an example of where e.g. a data race only shows up during the contention path of a mutex.
We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).
%
To ensure a fair comparison, we disabled \landslide's {\em false-positive}-avoidance techniques during this experiment.
For example, we reported free-re-malloc data races during the first interleaving as {\em deterministic}, as a single-pass analysis must, rather than waiting until future interleavings to confirm them (see Section \ref{sec:free-re-malloc}).

### 2015-10-10 pldi-16/related.tex
talk about dr pps

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist}, and ETA \cite{parallel-dpor}, require thread communication to be restricted to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is still applicable to these tools.

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
Ski \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
% TODO: Make sure you talk about your mutex experiment, or rephrase this.
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our mutex-testing experiment.
However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.

@@

% TODO: talk about data race detectors???

>>>>

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, \textsc{MaceMC} \cite{macemc}, \textsc{MoDist} \cite{modist}, and ETA \cite{dbug-retreat}, limits thread communication to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is still applicable to these tools.

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs. However, every such prior tool we know of has serious drawbacks.
{\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
% TODO: Make sure you talk about your mutex experiment, or rephrase this.
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our mutex-testing experiment.
However, it does not employ Iterative Deepening, and requires programs to be written in the PROMELA DSL, so cannot check implementations directly.

@@

% TODO: talk about data race detectors???
% eg Scalable Race Detection for Android Applications -- uses domain specific heuristics to filter out false positives
